### 2017-12-10 readme.rmd
added support landscape

@@

One of the most critical steps to take is to change the way we talk about the problem. In particular, insignificant phylogenetic results should not be presented as definitive resolutions of tough problems. This should always apply, but especially so in the case of highly contested questions of such broad interest. Making strong statements about insignificant results gives the impression that people with different interpretations of the available results are much further apart than they actually are. Making strong statements about insignificant results eventually gives the impression that phylogenetic methods are broken. They are not. How we talk about phylogenies is broken.

@@

When I look at the entirety of the impressive work done by multiple very talented groups on testing Ctenophora-sister and Porifera-sister, I come to the conclusion that at this time we cannot reject either hypothesis. I just don't find the sensitivity to species sampling, gene sampling, model selection, and now character coding conclusive. It isn't that I don't think it is impossible to come to a well-supported result if support for a phylogenetic hypothesis is sensitive to any of these factors, I just don't think we understand the sensitivity of this problem well enough to come to strong conclusions.

I also find the interactions between these factors concerning. I think that the rigorously blocked approach that Feuda *et al.* 2017 very exciting and the way of the future. It can provide a clear picture of the support landscape as it depends on different factors and their interactions. We are just looking at too small a part of the sensitivity landscape at too coarse a scale to understand the problem well enough yet.

I am optimistic that there will be significant progress on this problem in coming years. Analysis methods and tools are rapidly improving. New high-quality genome data will soon be available for many more species that are highly relevant to this question. For example, just a few days after Feuda *et al.* 2017 was published, [Belahbib *et al.*](https://www.biorxiv.org/content/early/2017/12/04/228452) posted a preprint presenting two new sponge genome sequences and analyses directly relevant to interpreting the biological significance of Ctenophora-sister and Porifera-sister. I look forward to to seeing how further advances in analyses and sampling impact tests of Ctenophora-sister and Porifera-sister and their impact on our understanding of animal evolution.

>>>>

## Using sensitivity to understand the support landscape

The sensitivity of phylogenetic analyses to taxon sampling, gene sampling, and analysis methods is often viewed as troublesome result. It provides very important insight, though, into the methods we use and the biological questions we apply them to. After ten years of intensive study, the placement of sponges and ctenophores have become one of the most thoroughly explored difficult phylogenetic problems. Examining the sensitivity of results will help us advance this particular question, and may also provide insight into how to address other hard phylogenetic questions.

Feuda *et al.* explain why they interpret the sensitivity they and others present as support for Porifera-sister in a few passages, including:

> The relationships at the root of the animal tree depend on model adequacy, with Ctenophora-sister emerging more prominently when the data are modeled less adequately and Porifera-sister being better supported when the data are more adequately modeled (Figure 1). As phylogenetic methods are less likely to recover artifactual topologies when the models of evolution more adequately describe the data [16–20], we conclude that Porifera-sister better represents the phylogenetic signal in the data.

> This model-dependent outcome has been interpreted as reflecting a fundamental lack of robustness [59], and it has been suggested that the relationships at the root of the animal tree might be impossible to reconstruct based on current amino acid datasets [2, 4]. However, sensitivity to the model (and to other analytical factors) does not imply an impossibility to decide which solution is most likely to be correct... the varied outcomes of phylogenetic analyses of animal relationships show a clear pattern, with Ctenophora-sister being systematically associated with the use of inadequate models and/or taxon-sampling schemes that are most likely to exacerbate systematic errors.

It is certainly true that model sensitivity alone does not mean we can reject a given phylogenetic hypothesis. There are an infinite number of possible models, and it is essentially certain that for any given phylogenetic hypothesis there is a model out there somewhere that will reject it. The question is how we use variation in support to put each result in a broader context that allows us to interpret it.

We can think of this variation as a support landscape - a multidimensional space that shows how support varies across different gene sampling, taxon sampling, and analysis methods. It is only ever possible to probe a finite region of this landscape, but its overall shape has radical implications for how we interpret a particular set of results. This means that when investigating sensitivity it is necessary to be explicit not only about the results we get in a given study, but how we are thinking about regions of the support landscape that are adjacent to those we examine.

The excerpts from Feuda *et al.* suggest that the authors implicitly envision the broader support landscape to be a monotonic surface (Figure 4a). They see believe that the small part of the support landscape they probed (Figure 4, inset boxes) indicates a broader pattern that they interpret as strong support for Porifera-sister. This is because, although most analyses they did support Ctenophora-sister there are some analyses on the right edge of the sampled space that recover significant support for Porifera-sister.

![overview](figures/figure_4.png)
*Figure 4 - Hypothetical support landscapes. The inset box shows the region of the landscape explored by Figures 2-3. The horizontal axis represents variation in analysis methods, such as model of molecular evolution and recoding of data. The vertical axis represents variation in sampling of genes and species. (a) A Monotonic Support Landscape. (b) A Sponge-Island Support Landscape, which would indicate that sponges support is only obtained in a very small portion of the support landscape. (c) A Hilly Support Landscape.*

Other support landscapes would have very different implications for the results they obtained. A small fraction of analyses may have supported Porifera-sister because they happened to sample a small island of support (Figure 4b) in a broader landscape that overall his patterns of support consistent with Ctenophora-sister. Alternatively, the small island of support for Porifera-sister may indicate a transition from a plateau of support for Ctenophora sister to a hilly support landscape to the east. It is unlikely that we would conclusively resolve the placement of ctenophores and sponges if this broader hilly pattern were found to be the case.

The work of Feuda *et al.* 2017 provides an enticing peek at a portion of this landscape, but without broader context still fails to allow us to come to strong conclusions. Exploring additional sensitivity will help us understand this bigger picture. In this sense, sensitivity is an asset rather than a liability.

@@

One of the most critical steps to take is to change the way we talk about the problem. In particular, insignificant phylogenetic results should not be presented as definitive resolutions of tough problems. This should always apply, but especially so in the case of highly contested questions of such broad interest. Making strong statements about insignificant results gives the impression that phylogenetic methods are broken as a question seems to bounce from one extreme answer to the other. They are not. How we talk about phylogenies is broken.

@@

When I look at the entirety of the impressive work done by multiple very talented groups on testing Ctenophora-sister and Porifera-sister, I come to the conclusion that at this time we cannot reject either hypothesis. I just don't find the sensitivity to species sampling, gene sampling, model selection, and now character coding conclusive. This is because I think we need a broader perspective of the support landscape to put the interesting results we already have into context. It isn't that I don't think it is impossible to come to a well-supported result if support for a phylogenetic hypothesis is sensitive to any of these factors, I just don't think we understand the sensitivity of this problem well enough to come to strong conclusions at this time.

I also find the interactions between these factors concerning. I think that the rigorously blocked approach that Feuda *et al.* 2017 very exciting and the way of the future. It can provide a clear picture of the support landscape as it depends on different factors and their interactions. We need to expand this approach.

I am optimistic that there will be significant progress on this problem in coming years. Analysis methods and tools are rapidly improving. New high-quality genome data will soon be available for many more species that are highly relevant to this question. For example, just a few days after Feuda *et al.* 2017 was published, [Belahbib *et al.*](https://www.biorxiv.org/content/early/2017/12/04/228452) posted a preprint presenting two new sponge genome sequences and analyses directly relevant to interpreting the biological significance of Ctenophora-sister and Porifera-sister. I look forward to to seeing how further advances in analyses and sampling push tests of Ctenophora-sister and Porifera-sister into new areas of the support landscape and allow us to chart its broader features. And, of course, I am excited about the impact this work will have on our understanding of animal evolution.
